hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-03-03 inst/manuscript/paper.Rmd
Introduction edits

This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.

@@

To maximize accessibility, **hatchR** is available in two formats. The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes. Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>). The second is a Shiny-based web application (@chang2024), offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality. Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.

>>>>

This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.

@@

To maximize accessibility, **hatchR** is available in two formats. The first is a R package, **hatchR**, distributed via CRAN [@r:ther], providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes. Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>). The second is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality. Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.
